---
title: "How Severe Weather Impacts us"
output:
  html_document:
    df_print: paged
    theme: spacelab
    highlight: pygments
    number_sections: yes
  html_notebook: default
  pdf_document: 
    highlight: pygments
    number_sections: yes
---
*Coursera - Reproducible Research - Project 2*

# Synopsis 
This analysis is for the Coursera Reproducible Research course Week 4 Project 2, part of the Data Science Specialization. The project is to explore the impact of severe weather events on human heatlth (injuries and 
fatalities) and finanical damage (property and crop damage)

# Data Processing

```{r, message=FALSE}
library(tidyverse)
library(grid)
library(gridExtra)
library(pander)
```

```{r Preprocessing, message=FALSE,cache=TRUE}
NOAA <- read_csv("repdata-data-StormData.csv")

names(NOAA)

stormData <- NOAA[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG",
                     "PROPDMGEXP","CROPDMG","CROPDMGEXP")]

```

## Data Processing for Health Analysis

```{r Health Analysis}
Health <- stormData %>% group_by(EVTYPE) %>% summarise(Injury = sum(INJURIES), Fatality =sum(FATALITIES))

HealthInjury <- Health [order (Health$Injury, decreasing = T), -3]

HealthFatality <- Health [order (Health$Fatality, decreasing = T), -2]
```

## Data Processing for Economic Damage Analysis



```{r , message=FALSE,warning=FALSE}
# check PROPDMGEXP class

class(stormData$PROPDMGEXP)

```

```{r}
# convert to factor

stormData$PROPDMGEXP <- as.factor(stormData$PROPDMGEXP)

stormData$CROPDMGEXP <- as.factor(stormData$CROPDMGEXP)

# subset stormData

ED <- stormData[,-c(2,3)]

# check PROPDMGEXP levels

levels(ED$PROPDMGEXP)
```


```{r}
table(factor(ED$PROPDMGEXP))

sum(is.na(ED$PROPDMGEXP))

```

```{r , message=FALSE}
ED$Prop.DMG[ED$PROPDMGEXP %in% c("-","?","+")] <- 0

ED$Prop.DMG[ED$PROPDMGEXP %in% c("h","H")] <- 100

ED$Prop.DMG[ED$PROPDMGEXP == "0"] <- 1

ED$Prop.DMG[ED$PROPDMGEXP == ""] <- 0

ED$Prop.DMG[ED$PROPDMGEXP == "1"] <- 10

ED$Prop.DMG[ED$PROPDMGEXP == "2"] <- 100

ED$Prop.DMG[ED$PROPDMGEXP == "3"] <- 1000

ED$Prop.DMG[ED$PROPDMGEXP == "4"] <- 1*10^4

ED$Prop.DMG[ED$PROPDMGEXP == "5"] <- 1*10^5

ED$Prop.DMG[ED$PROPDMGEXP == "6"] <- 1*10^6

ED$Prop.DMG[ED$PROPDMGEXP == "7"] <- 1* 10^7

ED$Prop.DMG[ED$PROPDMGEXP == "8"] <- 1* 10^8

ED$Prop.DMG[ED$PROPDMGEXP == "K"] <- 1000

ED$Prop.DMG[ED$PROPDMGEXP == "B"] <- 1*10^9

ED$Prop.DMG[ED$PROPDMGEXP %in% c("m","M")] <- 1*10^6

sum(is.na(ED$Prop.DMG))

Property <- ED[, c(1,2,6)]     

```


```{r , message=FALSE}

levels(ED$CROPDMGEXP)

table(factor(ED$CROPDMGEXP))

ED$Crop.DMG[ED$CROPDMGEXP == "?"] <- 0

ED$Crop.DMG[ED$CROPDMGEXP == ""] <- 0

ED$Crop.DMG[ED$CROPDMGEXP == "0"] <- 0

ED$Crop.DMG[ED$CROPDMGEXP == "2"] <- 100

ED$Crop.DMG[ED$CROPDMGEXP == "B"] <- 1*10^9

ED$Crop.DMG[ED$CROPDMGEXP %in% c("k","K")] <- 1000

ED$Crop.DMG[ED$CROPDMGEXP %in% c("m","M")] <- 1*10^6

Crop <- ED[,c(1,4,7)]

```


# Results

## Health Impact

Top 6 weather events cause injuries.
```{r Health Injury}

pander(head(HealthInjury,6))
```

Top 6 weather events cause fatalities.
```{r Health Fatality}

pander(head(HealthFatality,6))
```


## Economic Damage

```{r}
Property$DMG.Value <- Property$PROPDMG*Property$Prop.DMG

Crop$DMG.Value <- Crop$CROPDMG*Crop$Crop.DMG

PropertyDMD <- Property %>% group_by(EVTYPE) %>% 
    summarise(Damage = sum(DMG.Value)) %>% 
    arrange(desc(Damage))

PropertyDMDTop5 <- PropertyDMD [c(1:5),]

PropertyDMDTop5$Damage <- PropertyDMDTop5$Damage/10^9

names(PropertyDMDTop5) <- c("Events","Billions")

CropDMD <- Crop %>% group_by(EVTYPE) %>% 
    summarise(Damage = sum(DMG.Value)) %>% 
    arrange(desc(Damage))

CropDMDTop5 <- CropDMD[c(1:5),]

CropDMDTop5$Damage <- CropDMDTop5$Damage/10^9

names(CropDMDTop5) <- c("Events","Billions")

pander(PropertyDMDTop5)

pander(CropDMDTop5)

```

Top 5 weather events on Economic Impact

```{r , fig.width=9 }
g1 <- ggplot(PropertyDMDTop5,aes(x=Events, y = Billions)) + geom_col()+
    labs(title = "Property Damage (Top 5)")

g2 <- ggplot(CropDMDTop5,aes(x=Events, y = Billions)) + geom_col()+
    labs(title = " Crop Damage (Top 5)")

grid.arrange(g1,g2,nrow = 2)

```

